Flabby and injective objects in toposes

### 2018-06-20 paper-flabby-objects.tex
Flabby objects in the effective topos

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools,tikz,float,caption}
\usetikzlibrary{patterns,matrix,decorations.pathreplacing}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{hyperref}

@@

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}

@@

@@

\item In the ef{}fective topos, the statement ``if~$X$ is flabby, then any map~$X
\to X$ has a fixed point'' holds for ef{}fective sets~$X$. In particular, the
only module in the ef{}fective topos which is flabby and an ef{}fective set is the
trivial module (XXX proof). On the other hand, assuming the law of excluded middle, the
statement ``any module with~$\neg\neg$-stable equality can be embedded into a
flabby module'' holds in the ef{}fective topos. This is because for such a
module~$M$, the canonical morphism~$M \to \Delta(\Gamma(M))$ is a monomorphism,
the set~$\Gamma(M)$ is flabby by virtue of being inhabited, and~$\Delta$
preserves flabby objects by Proposition~\ref{prop:pushforward-of-flabby-objects}.
Here~$(\Delta \dashv \Gamma) : \Set \to \Eff$ is the inclusion of the
double-negation sheaves.
\end{enumerate}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools,tikz,float,caption,stmaryrd}
\usetikzlibrary{patterns,matrix,decorations.pathreplacing}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{hyperref}

@@

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}

@@

\section{Flabby objects in the effective topos}

The notion of flabby objects originates from the notion of flabby shaves and is
therefore closely connected to Grothendieck toposes. It is therefore instructive
to study flabby objects in elementary toposes which are not Grothendieck
toposes, away from their original conceptual home. We begin this study with
establishing the following observations on flabby objects in the effective
topos.

\begin{prop}\label{prop:flabby-effective-sets}
Let~$X$ be a flabby object in the effective topos. Let~$f : X \to X$
be a morphism. If~$X$ is effective, the statement ``$f$ has a fixed point''
holds in the effective topos.
\end{prop}

\begin{prop}\label{prop:semienough-flabby-modules}
Assuming the law of excluded middle, any~$\neg\neg$-separated
module in the effective topos can be embedded into a flabby module.
\end{prop}

We follow the terminology of~\cite{hyland:effective-topos}(XXX). The intuitive
reason for why Proposition~\ref{prop:flabby-effective-sets} holds is the
following. Let~$X$ be a flabby object of the effective topos. Then there is a
procedure which computes for any subterminal~$K \subseteq X$ an element~$x_K$
such that~$K \subseteq \{ x_K \}$. This element might not depend extensionally
on~$K$, but this fine point is not important for the further discussion. Let
now~$f : X \to X$ be a morphism. We construct the self-referential subset~$K
\defeq \{ f(x_K) \}$; the formal proof below will indicate how this can be
done. Then~$K \subseteq \{ x_K \}$, so~$f(x_K) = x_K$.

A corollary of Proposition~\ref{prop:flabby-effective-sets} is that the trivial
module is the only flabby module in the effective topos whose underlying
unstructured object is an effective set: Given such a flabby module~$M$, let~$v
\in M$ be an arbitrary element. Then the morphism~$x \mapsto v + x$ has a fixed
point; thus~$v + x = x$ for some element~$x$, and hence~$v = 0$.

It is the self-referentiality which makes the proof of
Proposition~\ref{prop:flabby-effective-sets} work, but the blame for paucity
of flabby objects in the effective topos is to put on the realizers for
statements of the form~``$K = K$'', where~$({=})$ is the nonstandard equality
predicate of the powerobject~$\P(X)$. A procedure witnessing flabbiness has to
compute a reflexivity realizer for a suitable element~$x_K$ from a reflexivity
realizer for a given element~$K$. However, such realizers are not very
informative. Metaphorically speaking, a procedure witnessing flabbiness has to
conjure elements out of thin air.

This problem does not manifest with objects~$X$ which are not effective sets.
Reflexivity realizers for these objects may not be very informative themselves;
a procedure witnessing flabbiness therefore only has to turn one kind of
non-informative realizers into another kind. The flabby modules featuring in
the proof of Proposition~\ref{prop:semienough-flabby-modules} will accordingly
be not effective sets.

\begin{proof}[Proof of Proposition~\ref{prop:flabby-effective-sets}]
For any Turing machine~$e$, let~$v_e$ be the set
\begin{multline*}
  v_e = \{ m \in \NN \,|\,
  \text{there is an element~$x_0 \in |X|$ such that} \\
  \text{$e$ terminates with an element of~$\llbracket x_0 = x_0 \rrbracket$ and
  $m \in \llbracket x = x_0 \rrbracket$} \}
\end{multline*}
and let~$K_e \in \Sigma^{|X| \times \Sigma}$ be the nonstandard predicate given by
\[ K_e(x,u) = \llbracket (x = x) \wedge (u \leftrightarrow v_e) \rrbracket. \]
One can explicitly construct a realizer~$a_e$ of the statement~``$K_e = K_e$'',
where~$({=})$ is the nonstandard equality predicate of the object~$\P_{\leq1}(X)$
of subterminals of~$X$. This is where the assumption that~$X$ is effective is
important; without it, we could only verify~``$K_e = K_e$'' where~$({=})$ is
the nonstandard equality predicate of the full powerobject~$\P(X)$.

Since~$X$ is flabby, there is a realizer~$r$ for the statement~``$\forall K \in
\P_{\leq1}(X)\_ \exists x \in X\_ \forall y \in X\_ (y \in K \Rightarrow y =
x)$''. Let~$s$ be a realizer for the statement~``$\forall x \in X\_ \exists y
\in Y\_ y = f(x)$''. Let~$e$ be the particular Turing machine which proceeds as
follows:
\begin{enumerate}
\item[1.] Simulate~$r$ on input~$a_e$ in order to obtain a realizer~$b \in
\llbracket x = x \rrbracket$ for some~$x \in |X|$.
\item[2.] Simulate~$s$ on input~$b$ in order to obtain a realizer~$c \in \llbracket
f(x) = f(x) \rrbracket$.
\item[3.] Output~$c$.
\end{enumerate}
The description of the machine~$e$ makes use of the number~$e$ coding it;
the recursion theorem yields a general reason why this self-referentiality is
possible. Here we can even do without this theorem, since a close inspection of
the construction of~$a_e$ shows that~$a_e$ is actually independent of~$e$. This
should not come as a surprise, as reflexivity realizers of~$\P(X)$
and~$\P_{\leq1}(X)$ are known to be not very informative.

Passing~$a_e$ to~$r$ yields a reflexivity realizer of some element~$x_{K_e} \in
|X|$. Therefore the Turing machine~$e$ does terminate, with a reflexivity
realizer for~$f(x_{K_e})$. Thus the statement~``$f(x_{K_e}) \in K$'' is
realized; hence~``$f(x_{K_e}) = x_{K_e}$'' is as well.
\end{proof}

\begin{proof}[Proof of Proposition~\ref{prop:semienough-flabby-modules}]
Let~$(\Delta \dashv \Gamma) : \Set \to \Eff$ be the inclusion of the
double-negation sheaves. For a~$\neg\neg$-separated module~$M$ in the effective
topos, the canonical morphism~$M \to \Delta(\Gamma(M))$ is a monomorphism; the
set~$\Gamma(M)$ is flabby by virtue of being inhabited; and~$\Delta$ preserves
flabby objects by Proposition~\ref{prop:pushforward-of-flabby-objects}.
\end{proof}

@@

\item There appears to be some tension regarding the effective topos:
Proposition~\ref{prop:semienough-flabby-modules} shows that at
least~$\neg\neg$-separated modules in the effective topos always embed into
flabby modules, assuming the law of excluded middle, while
Proposition~\ref{prop:flabby-effective-sets} shows that no nontrivial effective
module is flabby.
\end{enumerate}
